---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Akosua Gyan Bandoh"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#checking working directory
getwd()


#loading packages
library(tidyverse)
library(lubridate)
library(here)

#importing data
NTL_LTER_raw_data <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)

#2 setting theme

mytheme <- theme_classic(base_size = 12) + theme(axis.text.x = element_text(colour = "blue"),
                                                 axis.title.x = element_text(size = 12),
                                                 axis.title.y = element_text(size=12),
                                                 legend.position = "left",
                                                 plot.title = element_text(hjust = 0.5, size = 12))
theme_set(mytheme)
                                            



```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no significant difference in mean lake temperature across different depths during July.
Ha: There is a significant difference in mean lake temperature across different depths during July.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
# format date
NTL_LTER_raw_data$sampledate <- as.Date(NTL_LTER_raw_data$sampledate, format ="%m/%d/%y")

NTL_July <- NTL_LTER_raw_data %>% 
  filter(month(sampledate)==7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
 na.omit()


#5
# Create scatter plot with linear regression line
temperature_by_depth <- ggplot(NTL_July, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5, color = "blue") + 
  geom_smooth(method = "lm", col = "red", se = TRUE) +  
  labs(
    x = "Depth (m)", 
    y = "Temperature (°C)", 
    title = "Temperature by Depth in July"
  ) +
  ylim(0, 35)  # Limit temperature range

# Display the plot 
temperature_by_depth


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: There is an inverse relationship between temperature and depth. The distribution does not suggest linearity especially when the depth goes beyond 5m.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
linear_regression <- lm(temperature_C ~ depth, data = NTL_July)
summary(linear_regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The model suggests that for every 1-meter increase in depth, temperature decreases by approximately 1.95°C. The intercept of 21.96°C represents the predicted surface temperature at 0 meters depth. With 9,726 degrees of freedom, the model is based on a large dataset, making the findings robust. The p-value is less than 2.2e-16, meaning the relationship between depth and temperature is highly significant and unlikely to be due to random chance. Additionally, the R-squared value of 0.7387 indicates that 73.87% of the variation in temperature can be explained by depth alone, confirming that depth is a strong predictor of lake temperature.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
mlt.linear_regression <- lm(data = NTL_July, temperature_C ~ year4 + daynum + depth)

step(mlt.linear_regression)
#10
summary(mlt.linear_regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 

The AIC-selected model includes year, day of the year, and depth as explanatory variables for predicting lake temperature. This model explains 74.12% of the variance, slightly improving over the depth-only model (73.87%). While depth remains the strongest predictor, adding seasonal (daynum) and long-term (year) trends provides a more comprehensive understanding of temperature variation

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#aov
NTl_july_anova1 <- aov(data = NTL_July, temperature_C ~ lakename)
summary(NTl_july_anova1)

#lm
NTl_july_anova2 <- lm(data = NTL_July, temperature_C ~ lakename)
summary(NTl_july_anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 

The ANOVA test (F = 50.0, p < 2.2e-16) shows a highly significant difference in mean temperature, meaning that lake temperature varies significantly across different lakes. Hence we can reject the null hypothesis (H₀) that there is no significant difference in mean temperature among the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(NTL_July, aes(x=depth, y= temperature_C, color = lakename)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_smooth(method = "lm" , se = F) +
  labs(x= "Depth (m)", y = "Temperature (C)", 
       title = "Temperature by Depth in July", color = "lakename") +
  ylim(0,35)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(NTl_july_anova1)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
The Tukey HSD test results indicate that Peter Lake has statistically similar mean temperatures to Paul Lake and Ward Lake, as their p-values are greater than 0.05 and their confidence intervals include zero. This means we cannot conclude that their temperatures are significantly different. On the other hand, East Long Lake stands out as having a statistically distinct mean temperature from multiple lakes, suggesting it is significantly colder compared to others. While some lakes share similar temperature patterns, others, like East Long Lake, Tuesday Lake, and West Long Lake, exhibit clear temperature differences from the rest.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were just comparing Peter Lake and Paul Lake, a T-test would be a suitable method to determine whether their mean temperatures are significantly different.A two-sample t-test (independent t-test) would help compare their means.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
July_Crampton_Ward <- NTL_July %>% 
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

#test
two_sample_test <-t.test(July_Crampton_Ward$temperature_C ~ July_Crampton_Ward$lakename)
two_sample_test
```

>Answer: According to the test, the p-value is 0.2649, greater than 0.05. Meaning the mean temperatures for Crampton and Ward lakes are statistically the same. This matches the answer in part 16.
