---
title: "Assignment 8: Time Series Analysis"
author: "Akosua Bandoh"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}
library(tidyverse)
library(lubridate)
install.packages("trend")
install.packages("trend")
library(trend)
#install.packages("zoo")
library(zoo)
install.packages("Kendall")
library(Kendall)
#install.packages("tseries")
library(tseries)
library(here)

#checking working directory
here()

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#2 Importing datasets

#all datasets
all_datasets <- list.files ("Data/Raw/Ozone_TimeSeries/", pattern = "*.csv",
                           full.names = TRUE)
GaringerOzone <- lapply (all_datasets,read.csv, stringsAsFactors = TRUE) %>% 
  bind_rows()


```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3
GaringerOzone$Date <- as.Date(GaringerOzone$Date, format = '%m/%d/%Y')

# 4
GaringerOzone <- select(GaringerOzone, Date, 
                            Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE)

# 5
Days <- as.data.frame(seq(as.Date("2010-01-01"), as.Date("2019-12-31"), by = 'day'))
#rename column nae in Days to Date
names(Days) <- 'Date'

# 6
GaringerOzone <- Days %>% 
  left_join(GaringerOzone, by = "Date")

#checking dimensions
dim(GaringerOzone)



```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7
ggplot(GaringerOzone, aes(x = Date, y = Daily.Max.8.hour.Ozone.Concentration)) +
  geom_line() + 
  geom_smooth(method = 'lm') + scale_x_date(date_breaks = '1 year', date_labels = '%b %Y') +
  labs(
    title = "Ozone Concentrations", 
    x = "Date",
    y = "Daily Concentration (ppm)"
  )

```

>Answer:There is a slight decrease in ozone concentration over time

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8
GaringerOzone_Fill <- GaringerOzone %>% 
  mutate(Daily.Max.8.hour.Ozone.Concentration = 
           zoo::na.approx(Daily.Max.8.hour.Ozone.Concentration))

#checking for NAs
summary(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration)

```

> Answer: We didn’t use piecewise constant interpolation because it would just repeat the last known value until the next data point, which isn’t realistic since ozone levels change daily. We avoided spline interpolation because it can create weird curves and swings that don’t actually happen in real life.


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9
GaringerOzone.monthly <- GaringerOzone_Fill %>% 
  mutate(
    Month = month(Date), Year = year(Date)
  ) %>% 
  group_by(Year, Month) %>% 
  summarise(MonthlyMeanOzone = mean(Daily.Max.8.hour.Ozone.Concentration))

GaringerOzone.monthly <- GaringerOzone.monthly %>% 
  mutate(Date = paste(Year, Month, "01", sep = "-"))

```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10
GaringerOzone.daily.ts <-
  ts(GaringerOzone_Fill$Daily.Max.8.hour.Ozone.Concentration, 
                             start = c(2010,1), frequency = 365)

GaringerOzone.monthly.ts <-
  ts(GaringerOzone.monthly$MonthlyMeanOzone, start = c(2010,1), frequency = 12)
   
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11
Decomposed_GaringerOzone_Fill <- stl(GaringerOzone.daily.ts, s.window = 'periodic')

plot(Decomposed_GaringerOzone_Fill)


Decomposed_GaringerOzone_monthly_Fill <- stl(GaringerOzone.monthly.ts, s.window = 'periodic')

plot(Decomposed_GaringerOzone_monthly_Fill)

```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12
GaringerOzone.monthly_trend <- Kendall::SeasonalMannKendall(GaringerOzone.monthly.ts)

GaringerOzone.monthly_trend

summary(GaringerOzone.monthly_trend)

GaringerOzone.monthly_trend1 <- trend:: smk.test(GaringerOzone.monthly.ts)

GaringerOzone.monthly_trend1

```

> Answer: The Seasonal Mann-Kendall test is the best choice because it detects monotonic trends in time series data while accounting for seasonal patterns.

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
# 13
ggplot(GaringerOzone.monthly, aes(x = Date, y = MonthlyMeanOzone)) +
  geom_point() +  
  geom_line(color = 'red') + 
  scale_x_date(date_breaks = '1 year', date_labels = '%b %Y') +
  labs(
    title = "Ozone Concentrations",
    x = "Date",
    y = "Monthly Concentration (ppm)"  
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: The analysis of monthly ozone concentrations at Garinger High School from 2010 to 2019 reveals both a slight decreasing trend and a seasonal pattern in ozone levels. The Seasonal Mann-Kendall test confirms a statistically significant downward trend (tau = -0.143, p = 0.0467), suggesting a gradual decline in ozone concentrations over the decade. However, the graph shows recurring seasonal fluctuations, likely due to higher ozone levels in warmer months and lower levels in colder months, a common pattern influenced by temperature, sunlight, and atmospheric conditions.

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15

GaringerOzone.monthly.seasonal <- GaringerOzone.monthly.ts - 
  Decomposed_GaringerOzone_monthly_Fill$time.series[, "seasonal"]

GaringerOzone.monthly.seasonal.ts <- ts(GaringerOzone.monthly.seasonal,
                                        start = start(GaringerOzone.monthly.ts),
                                        frequency = frequency(GaringerOzone.monthly.ts))

#16
GaringerOzone.monthly.seasonal_trend <- Kendall::MannKendall(GaringerOzone.monthly.seasonal.ts)

GaringerOzone.monthly.seasonal_trend

summary(GaringerOzone.monthly.seasonal_trend)


```

> Answer: The p-value comparison shows that the trend becomes more statistically significant after removing seasonal effects. The Seasonal Mann-Kendall test detected a weak but significant decline (p = 0.0467), while the de-seasonalized test showed a stronger decreasing trend (p = 0.0075). This suggests that while seasonal fluctuations influenced ozone levels, the long-term decline is real. Likely drivers include air quality regulations, emission reductions, and environmental policies, indicating a sustained improvement in ozone pollution over the decade.
