---
title: "Assignment 3: Data Exploration"
author: "Akosua Gyan Bandoh"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
#Loading packages
library('tidyverse')
library('lubridate')
library('here')

#checking working directory
getwd()

#Loading datasets
Neonics <- read.csv(
  file = here('Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'),
  stringsAsFactors = TRUE
)

Litter <- read.csv(
  file = here('Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
  stringsAsFactors = TRUE
)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Neonicotinoids are insecticides that impact pollinators and beneficial insects, potentially affecting ecosystems and food production.By analyzing the data, we can assess the risks posed by neonicotinoids to different insect populations. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Litter and woody debris play a cricitical role in the carbon and nitrogen cycles within forest ecosystems. As they decompose, they release nutrients back into the soil, which supports plant growth and maintains soil fertility.By analyzing this data, we can gain insights into the functioning and sustainability of forest ecosystems, which is vital for both environmental health and biodiversity.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Litter and fine woody debris sampling is executed at terrestrial NEON sites that contain woody vegetatioon >2m tall.Sampling for this product occurs only in tower plots. Locations of tower plots are selected randomly within the 90% flux footprint of the primary and secondary airsheds.
 2.Trap placement within plots may be either targeted or randomized, depending on the vegetation. 
 3.Ground traps are sampled once per year.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
#Dimensions of the Neonics dataset
dim(Neonics)

#The dimensions are 4623 rows and 30 colums
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
sort(summary(Neonics$Effect))

```

> Answer:The most common effects studied are Population with a count of 1803 and Mortality with a count of 1493. This is of interest because they are great indicators of the impact of these insecticides on the insect population.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
#finding the six most commonly studied species
(summary(Neonics$Species.Common.Name, maxsum = 7))

```

> Answer:The six most most commonly studied species are all different types of bees - Honey Bee, Parasitic Wasp Buff Tailed BumbleBee, Carniolan honey Bee, Bumble Bee, Italian Honeybee. This is because bees are considered the best pollinators and their mortality will largely affect the ecosystem.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
#finding the class of Conc.1..Author
class(Neonics$Conc.1.Units..Author.)
```

> Answer: The class off conc.1.unit.. Authour is a factor. This is because the values in the column contain non-numeric characters. When the stringsAsfactors was used, R may have interpreted the non-numeric values and converted them to factors.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
#generating a plot for the number of studies conducted by publication year
ggplot(Neonics)+geom_freqpoly(aes(x = Publication.Year))

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x=Publication.Year,color = Test.Location))

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:The most common test location is the lab. Field natural is the next common location after the lab.The lab location increased after 2010 and significantly dropped after 2015. The field natural also dropped after 2010.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics,aes(x = Endpoint)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

> Answer: The most common Endpoints ae NEOL and LOEL. LOEL is defined as the Lowest Observed Effect Level and NOEL is defined as No Observed effect level. LOEL is the lowest concentration producing effects that were significantly different from the control. NOEL is the highest concentration of a substance where no statistically significant adverse effects are observed in a study.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
#Checking class of collectdate
class(Litter$collectDate)

#changing the date from factor
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")

#Checking class of collectdate
class(Litter$collectDate)

unique(Litter$collectDate)


```


```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)

summary(Litter$plotID)

```

> Answer:12 plots were sampled. The unique function determine how many different samples but does not give the frequency as the summary does.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter) + 
  geom_bar(aes(x=functionalGroup))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter) + 
  geom_boxplot(aes(x=functionalGroup, y= dryMass))

ggplot(Litter) + 
  geom_violin(aes(x=functionalGroup, y= dryMass))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: This is because there is fairly equal distribution across the Niwot Ridge sites. Vioilin plots are best for complex and skewed distributions.The boxplots showed the median, range and quartiles of the data.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles and Mixed
